\Title

### 2022-09-30 Grimaldi-etal-BiolCybernetics.tex
Update on Overleaf.

The challenges are numerous in such situations and two major obstacles must be highlighted: the computational delay and the energy budget. First, most existing solutions are built incrementally from models built to process low definition images, such as Le Net for MNIST~\citep{lecun1998gradient}, then high definition images, such as VGG on Imagenet~\citep{Simonyan:2014cmh}. This has a cost and notably the fact that the computation time increases with the size of the image. This phenomenon has been mitigated by the increase in available computing power, notably thanks to the parallel processing implemented in GPU cards. However, the transition from images to videos, then to multiple cameras implies a further level in the change of scale in the amount of data processed. It also raises an issue that is often not addressed when it comes to image categorization. Indeed, in autonomous driving, an important task is to react as quickly as possible to a possible collision, for example when a pedestrian unexpectedly crosses the street. As such, future systems should always be able to give a response at any time, instead of having to wait for the pipeline to provide the correct answer. This online processing is also a feature of processing in biological systems and could be an important first contribution if we could translate these principles into future automated processes. %
Second, a major constraint for future systems is energy cost. As we have seen, fast computation time has been maintained for high-dimensional images, but at the cost of higher energy consumption. In addition, the need to solve increasingly complex tasks has required the construction of increasingly complex and deep architectures. As an example, the best performing model in the Imagenet challenge uses 2100 million parameters. Training the model typically consumes kilowatts of power and about 1 joule to infer a decision per single image. Less power-hungry solutions have emerged, such as Mobile Net~\citep{howard_searching_2019}, but they are not yet ready for practical use in autonomous cars, as the additional power consumption would be a significant percentage of the energy consumed by an electric car under standard conditions.

@@

Neuroscience could provide revolutionary answers to these two closely related problems. Indeed, the human brain has the remarkable property of being able to react at any time, while consuming a reasonable amount of energy, about tens of watts. This system is the result of millions of years of natural selection, and a striking difference between biological neural networks and artificial ones is the representation they use. CNNs, for example, represent information passing from one layer to another as tensors representing visual information densely over the visual topography, while representing different properties in different channels. These networks have been known to mimick many propoerties of biological systems~\citep{Schrimpf2020}, yet, this score does not count rapidity or energy. On the other hand, information in a large majority of biological neural networks is represented as spikes, i.e. prototypical all-or-nothing events whose only parameters are their timing and the address of the neuron that emitted this spike. Why spikes? It is not yet known why most biological systems have evolved to use this representation, but one major advantage seems to be some form of efficiency. In the human brain, each of the approximately 90 trillion nerve cells emits on average one spike per second which fan out to an average of 10000 synapses. Importantly, each neuron fires with great variability, and up to 300 spikes per second. Many attempts have been made to build artificial spiking neural networks, and in particular to achieve the same performance as observed in classical CNNs with a lower energy budget. However, none of these networks has yet surpassed the state of the art performances. Here, we will try to focus on one core property of spiking neurons inspired by neurobiology to progress in this direction.

@@

\note{ TODO: formula / say it is driven by image gradients / give scaling rules }
To transform each movie into events, we compute a gradient image (initialized at zero) by adding the gradient of the pixels' intensity over two successive frames. If, on a specific pixel at that specific timestamp, the absolute value of this gradient exceeds a threshold, an event is generated. The event has either an OFF or ON polarity, respectively whether the gradient is negative or positive. This signed threshold value is then subtracted from the residual gradient image. When applied to the whole movie, the event stream is as a consequence similar to the output of a neuromorphic camera~\citep{rasetto_challenges_2022}, that is, a list of events defined by $x_\arank$ and $y_\arank$ (their position on the pixel grid), their polarity $\polev_\arank$ (ON or OFF) and time $\timev_\arank$  (\seeFig{motion_task}-(c)). The goal here is to infer the correct motion solely by observing these events. 
This sensory signal representing the output of this event-based camera forms a discrete stream of events, $\event$ as defined above.   
Each event has an associated address, which is typically in the form $\presynaddr_\arank=(x_\arank, y_\arank, \polev_\arank)$. This defines a presynaptic address space $\presynaddrspace = [1, \Nx]\times[1, \Ny] \times [1, \Npol] \subset \mathbb{N}^3$ where $(\Nx, \Ny)$ is the size of the sensor in pixels and $\Npol$ is the number of polarities  ($\Npol=2$ for ON and OFF polarities). 
\note{ the task is full-field rigid translations, the model resolves it locally (analogy to the input from V1 to MT) }
%
\subsection{Extension as 3D convolutions for motion detection}
%%%-----------------------------------------------------------------
%
Now that our motion detection task is defined in visual space, we can extend the convolution-based detection model to a 3D convolution so that the resulting model would also benefit from spatial invariance. The use of spatio-temporal filters on an event stream has improved the performance of CNNs for an action recognition task by~\citet{ghosh_spatio-temporal_2019} and here we will use a similar strategy, but focus on using an event-based representation.
For this purpose, we design 3D kernels of shape $(\Kx, \Ky, \Ktime) = (9, 9, 11)$, representing the two spatial dimensions and the range of time delays, respectively.
The computations are performed on the spatio-temporal windows that are defined by the kernels and are "moving" around the events, i.e. the spatial center of the spatio-temporal window around the current event $\epsilon_\arank$. Note that to remain within the framework of a causal calculation, the kernels are shifted in time such that only past information gives an answer at the present time.
%
The output of the MLR model corresponds to an event with the highest probability class, keeping the same timing as the input event. It is important to note that we include the prior knowledge that the event-based input is spatio-temporally clustered and that these clusters are sparse. Therefore, we will only select a given proportion of the most active cells, which we set here to $.01\%$ of all output voxels. This threshold was found using validation and future work should automate how this may be determined for various datasets. Given a stimulus input, the model thus yields a spike output in the postsynaptic address space.

%
Since the model is fully differentiable, we can now implement a semi-supervised learning rule. The loss function of the MLR model is the binary cross-entropy on the output of the classification layer and can be extended to space-time convolutions. Supervision was implemented using the input binary events as defined above and the labels as the desired output. The labels were defined at each time point as a one-shot encoding of the current motion in the channel corresponding to the current motion for all positions. As in this semi-supervised context, the label is known, but the timing is not, we used a selection of the temporal support in an unsupervised manner. It is important to note that we weighted the cost function by considering only active cells, such that the error is only backpropagated to the spatial locations of these most active cells. This is reminiscent of previous methods solving this problem using a Winner-Takes-All mechanism~\citep{masquelier2007}. Simulations are performed with the PyTorch library using gradient descent with Adam (for $2^{12}$ movies and a learning rate of $10^{-5}$). We tested the effect of different parameters on a validation set to quantify the role of each parameter.

@@

Let us highlight some innovations in the contributions presented in this paper. First, the generic heterogeneous model is formalized from first principles for optimal detection of the event-based spatio-temporal motifs, whereas~\citet{ghosh_spatio-temporal_2019} use a correlation-based heuristic, which we have observed to be less efficient. Moreover, in comparison to HOTS~\citep{lagorce_hots_2017} the weights are explainable as they directly inform on the logit (inverse sigmoid of the probability) of detecting each spatio-temporal spiking motif. Another novelty is that the model simultaneously learns the weights and the delays, while for example the polychronization model~\citep{izhikevich_polychronization_2006} only learns the weights using STDP while the delays are randomly drawn and their values frozen. Moreover, the model is evaluated on a realistic task, while models such as the tempotron are tested on simplified problems~\citep{gutig_tempotron_2006}. Another main contribution is to provide a model that is suitable for learning any type of spatio-temporal spiking motifs and that can be trained in a supervised way by providing a dataset of supervision pairs. This allows for a more flexible definition of the model using this properly labeled dataset instead of relying on a careful description of the physical rules governing a task, e.g., the luminance conservation principle for motion detection~\citep{benosman_asynchronous_2012, dardelet_event-by-event_2021}.
%
\subsection{Main limits}
%%%-----------------------------------------------------------------
We have identified several limits to our model, the most important we will now detail. First, the complete framework is based on a discrete binning of time that is incompatible with the continuous nature of biological time. We have used this binning to be able to efficiently implement the framework on conventional hardware, especially GPUs, and in particular to be able to use three-dimensional convolutions. We have tested the effect of the size of the time bin and shown that it has essentially no impact on the results presented in this paper. This is consistent with the relative robustness of other event-based frameworks such as HOTS~\citep{lagorce_hots_2017}, where the accuracy was not affected when the input spikes were subjected to noisy perturbations up to $1~\ms$~\citep{grimaldi_robust_2022}. This suggests the possibility of analytically including a precision term in the temporal value of the input spikes, a mechanism potentially implemented by the filtering that is implemented by the synaptic time constant of about $5~\ms$. Furthermore, it is possible to circonvent the necessity of using a time discretization by using a purely event-based scheme. Indeed, it is not necessary to compute voltage traces between two spikes~\citep{hanuschkin_general_2010} and it is thus possible to define a purely event-based framework. Such an architecture could provide promising speed gains for the calculations.  \note{pourrait parler du poster bernstein? de nadafian?}

>>>>

The challenges are numerous in such situations and two major obstacles must be highlighted: the computational delay and the energy budget. First, most existing solutions are built incrementally from models built to process low definition images, such as Le Net for MNIST~\citep{lecun_gradient-based_1998}, then high definition images, such as VGG on Imagenet~\citep{simonyan_very_2015}. This has a cost and notably the fact that the computation time increases with the size of the image. This phenomenon has been mitigated by the increase in available computing power, notably thanks to the parallel processing implemented in GPU cards. However, the transition from images to videos, then to multiple cameras implies a further level in the change of scale in the amount of data processed. It also raises an issue that is often not addressed when it comes to image categorization. Indeed, in autonomous driving, an important task is to react as quickly as possible to a possible collision, for example when a pedestrian unexpectedly crosses the street. As such, future systems should always be able to give a response at any time, instead of having to wait for the pipeline to provide the correct answer. This online processing is also a feature of processing in biological systems and could be an important first contribution if we could translate these principles into future automated processes. %
Second, a major constraint for future systems is energy cost. As we have seen, fast computation time has been maintained for high-dimensional images, but at the cost of higher energy consumption. In addition, the need to solve increasingly complex tasks has required the construction of increasingly complex and deep architectures. As an example, the best performing model in the Imagenet challenge uses 2100 million parameters. Training the model typically consumes kilowatts of power and about 1 joule to infer a decision per single image. Less power-hungry solutions have emerged, such as Mobile Net~\citep{howard_searching_2019}, but they are not yet ready for practical use in autonomous cars, as the additional power consumption would be a significant percentage of the energy consumed by an electric car under standard conditions.

@@

Neuroscience could provide revolutionary answers to these two closely related problems. Indeed, the human brain has the remarkable property of being able to react at any time, while consuming a reasonable amount of energy, about tens of watts. This system is the result of millions of years of natural selection, and a striking difference between biological neural networks and artificial ones is the representation they use. CNNs, for example, represent information passing from one layer to another as tensors representing visual information densely over the visual topography, while representing different properties in different channels. These networks have been known to mimick many propoerties of biological systems~\citep{schrimpf_brain-score_2020}, yet, this score does not count rapidity or energy. On the other hand, information in a large majority of biological neural networks is represented as spikes, i.e. prototypical all-or-nothing events whose only parameters are their timing and the address of the neuron that emitted this spike. Why spikes? It is not yet known why most biological systems have evolved to use this representation, but one major advantage seems to be some form of efficiency. In the human brain, each of the approximately 90 trillion nerve cells emits on average one spike per second which fan out to an average of 10000 synapses. Importantly, each neuron fires with great variability, and up to 300 spikes per second. Many attempts have been made to build artificial spiking neural networks, and in particular to achieve the same performance as observed in classical CNNs with a lower energy budget. However, none of these networks has yet surpassed the state of the art performances. Here, we will try to focus on one core property of spiking neurons inspired by neurobiology to progress in this direction.

@@

\note{ TODO: formula / say it is driven by image gradients / give scaling rules }
To transform each movie into events, we compute a gradient image (initialized at zero) by adding the gradient of the pixels' intensity over two successive frames. If, on a specific pixel at that specific timestamp, the absolute value of this gradient exceeds a threshold, an event is generated. The event has either an OFF or ON polarity, respectively whether the gradient is negative or positive. This signed threshold value is then subtracted from the residual gradient image. When applied to the whole movie, the event stream is as a consequence similar to the output of a neuromorphic camera~\citep{rasetto_challenges_2022}, that is, a list of events defined by $x_\arank$ and $y_\arank$ (their position on the pixel grid), their polarity $\polev_\arank$ (ON or OFF) and time $\timev_\arank$  (\seeFig{motion_task}-(c)). The goal here is to infer the correct motion solely by observing these events. 
This sensory signal representing the output of this event-based camera forms a discrete stream of events, $\event$ as defined above.   
Each event has an associated address, which is typically in the form $\presynaddr_\arank=(x_\arank, y_\arank, \polev_\arank)$. This defines a presynaptic address space $\presynaddrspace = [1, \Nx]\times[1, \Ny] \times [1, \Npol] \subset \mathbb{N}^3$ where $(\Nx, \Ny)$ is the size of the sensor in pixels and $\Npol$ is the number of polarities  ($\Npol=2$ for ON and OFF polarities). 
\note{ the task is full-field rigid translations, the model resolves it locally (analogy to the input from V1 to MT) }
%
\subsection{Extension as 3D convolutions for motion detection}
%%%-----------------------------------------------------------------
%
Now that our motion detection task is defined in visual space, we can extend the convolution-based detection model to a 3D convolution so that the resulting model would also benefit from spatial invariance. The use of spatio-temporal filters on an event stream has improved the performance of CNNs for an action recognition task by~\citet{ghosh_spatiotemporal_2019} and here we will use a similar strategy, but focus on using an event-based representation.
For this purpose, we design 3D kernels of shape $(\Kx, \Ky, \Ktime) = (9, 9, 11)$, representing the two spatial dimensions and the range of time delays, respectively.
The computations are performed on the spatio-temporal windows that are defined by the kernels and are "moving" around the events, i.e. the spatial center of the spatio-temporal window around the current event $\epsilon_\arank$. Note that to remain within the framework of a causal calculation, the kernels are shifted in time such that only past information gives an answer at the present time.
%
The output of the MLR model corresponds to an event with the highest probability class, keeping the same timing as the input event. It is important to note that we include the prior knowledge that the event-based input is spatio-temporally clustered and that these clusters are sparse. Therefore, we will only select a given proportion of the most active cells, which we set here to $.01\%$ of all output voxels. This threshold was found using validation and future work should automate how this may be determined for various datasets. Given a stimulus input, the model thus yields a spike output in the postsynaptic address space.

%
Since the model is fully differentiable, we can now implement a semi-supervised learning rule. The loss function of the MLR model is the binary cross-entropy on the output of the classification layer and can be extended to space-time convolutions. Supervision was implemented using the input binary events as defined above and the labels as the desired output. The labels were defined at each time point as a one-shot encoding of the current motion in the channel corresponding to the current motion for all positions. As in this semi-supervised context, the label is known, but the timing is not, we used a selection of the temporal support in an unsupervised manner. It is important to note that we weighted the cost function by considering only active cells, such that the error is only backpropagated to the spatial locations of these most active cells. This is reminiscent of previous methods solving this problem using a Winner-Takes-All mechanism~\citep{masquelier_unsupervised_2007}. Simulations are performed with the PyTorch library using gradient descent with Adam (for $2^{12}$ movies and a learning rate of $10^{-5}$). We tested the effect of different parameters on a validation set to quantify the role of each parameter.

@@

Let us highlight some innovations in the contributions presented in this paper. First, the generic heterogeneous model is formalized from first principles for optimal detection of the event-based spatio-temporal motifs, whereas~\citet{ghosh_spatiotemporal_2019} use a correlation-based heuristic, which we have observed to be less efficient. Moreover, in comparison to HOTS~\citep{lagorce_hots_2017} the weights are explainable as they directly inform on the logit (inverse sigmoid of the probability) of detecting each spatio-temporal spiking motif. Another novelty is that the model simultaneously learns the weights and the delays, while for example the polychronization model~\citep{izhikevich_polychronization_2006} only learns the weights using STDP while the delays are randomly drawn and their values frozen. Moreover, the model is evaluated on a realistic task, while models such as the tempotron are tested on simplified problems~\citep{gutig_tempotron_2006}. Another main contribution is to provide a model that is suitable for learning any type of spatio-temporal spiking motifs and that can be trained in a supervised way by providing a dataset of supervision pairs. This allows for a more flexible definition of the model using this properly labeled dataset instead of relying on a careful description of the physical rules governing a task, e.g., the luminance conservation principle for motion detection~\citep{benosman_asynchronous_2012, dardelet_event-by-event_2021}.
%
\subsection{Main limits}
%%%-----------------------------------------------------------------
We have identified several limits to our model which we will now detail. First, the complete framework is based on a discrete binning of time that is incompatible with the continuous nature of biological time. We have used this binning to be able to efficiently implement the framework on conventional hardware, especially GPUs, and in particular to be able to use three-dimensional convolutions. We have tested the effect of the size of the time bin and shown that it has essentially no impact on the results presented in this paper. This is consistent with the relative robustness of other event-based frameworks such as HOTS~\citep{lagorce_hots_2017}, where the accuracy was not affected when the input spikes were subjected to noisy perturbations up to $1~\ms$~\citep{grimaldi_robust_2022}. This suggests the possibility of analytically including a precision term in the temporal value of the input spikes, a mechanism potentially implemented by the filtering that is implemented by the synaptic time constant of about $5~\ms$. Furthermore, it is possible to circonvent the necessity of using a time discretization by using a purely event-based scheme. Indeed, it is not necessary to compute voltage traces between two spikes~\citep{hanuschkin_general_2010} and it is thus possible to define a purely event-based framework. Such an architecture could provide promising speed gains for the calculations.  \note{pourrait parler du poster bernstein? de nadafian?}
